% chapters/03_sections/baseline_selection
\section{Baseline Selection}\label{sec:baseline}

To make performance claims meaningful, all custom kernels are compared against
well-defined baselines under matched problem sizes, precision, and memory
layout assumptions.

\subsection{Primary Baselines}

\begin{itemize}
  \item \textbf{cuBLAS:} reference for GEMM-like contraction kernels.
  \item \textbf{cuTENSOR:} reference for direct tensor contraction and
    permutation-centric workflows.
  \item \textbf{cuBLASDx compositions:} reference for device-side small-GEMM
    execution where launch amortisation is central.
\end{itemize}

\subsection{Fairness Criteria}

Each comparison uses:

\begin{enumerate}
  \item identical input tensors and dimensions,
  \item identical precision modes and accumulation rules,
  \item warm-up runs and repeated measurements for variance control,
  \item consistent stream and synchronisation semantics.
\end{enumerate}

Reported metrics include median runtime, throughput (GFLOP/s or TFLOP/s),
effective memory bandwidth, and speedup relative to the strongest baseline per
workload.

\subsection{Profiling Evidence Gate}
\label{sec:profiling-evidence-gate}

To keep optimisation claims defensible, each accepted speedup is paired with a
matching profiler signal:

\begin{enumerate}
  \item \textbf{Launch-amortisation claim:} lower launch count and lower host
    or launch-overhead fraction at similar numerical output.
  \item \textbf{Memory-efficiency claim:} higher effective throughput and/or
    improved coalescing/reuse indicators with matching or lower runtime.
  \item \textbf{Occupancy/resource claim:} improved runtime with a coherent
    change in achieved occupancy and stall breakdown.
  \item \textbf{Precision-path claim (FP32/TF32):} runtime gain plus explicit
    numerical-deviation check against the FP64/FP32 reference path.
\end{enumerate}

This gate is applied uniformly to custom kernels and library-based baselines,
so comparison quality is maintained while the kernel set is still evolving.
